### 2016-04-27 hans/bayes-tardis-luminosity-prediction.tex
[draft] add hedgehog plot
[draft] hedgehog caption
[draft] Add missing N
[bayes] run dos2unix again
[draft] Clarify approximations

The Gamma distribution  with shape parameter $\alpha$ and rate
parameter $\beta$ is defined as
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak close to but not at the boundary. \fred{refer to plot of real
  luminosity packets in some frequency bin} In order to describe the
luminosity distribution of real packets through $\GammaDist(\alpha,
\beta)$, we have to transform the luminosities. To this end, we first
take the largest luminosity of all TARDIS packets across the entire
frequency range and increase it slightly to avoid numerical
difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.  We assume the total number of samples
across all bins is large, hence the effect of considering $\lmax$ a
constant independent of the data is negligible. We then transform the
samples $\bml \to \bmx$ as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond \nu,  \npack, \bmx,
\bmnu)$. Only in the end, we transform back to a probability density
in $L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax}
\end{align}
\hans{So far, $m$ appears only in your todo comment above. I would
  suggest a short discussion of this already in Section
  \ref{sec:dfnr}. We could then simply rewrite all equations below in
  terms of $m$ rather than $b$.} %
\fred{I removed $m$ whereever I saw it except below where I define it
  explicitly}
%
We observe good agreement between the best fit of a Gamma distribution
to TARDIS packet luminosities; an example is shown shown in
\reffig{gamma}.  While not obvious at first, a major numerical
simplification arises because of the $L \to X$
transformation. Rewriting the master equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
\nu, \npack, \bml, \bmnu)
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single luminosity, $p(L \cond N=1,
\nu, \bml, \bmnu)$ is narrow in the sense that the standard deviation
relative to the mean is at the percent level. This implies that $p(L
\cond \nu, \npack, \bml, \bmnu)$ has one mode from each $N$ and all
modes are clearly separated. The transformation $L \to x$ effects a
strong overlap between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
approximated by a unimodal distribution; see \reffig{hedgehog}.
\fred{I'm not 100 \% sure but $p(X \cond \nu, \npack, \bmx, \bmnu)$ should still
  have $N \to \infty$ modes but they aren't visible}
\missingfigure{from hedgehog to single peak \label{fig:hedgehog}}

@@

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
\end{align}
Let us assume that $\alpha$ and $\beta$ are functions of $\nu$ with unknown parameters $\bmphi$
\begin{align}
  \lleq{gmi}
  \alpha &= \alpha(\nu \cond \bmphi),\\
  \beta &= \beta(\nu \cond \bmphi),\\
  p(X_j \cond \bmphi, \nu) &=
  \GammaDist(X_j \cond \alpha, \beta),\\
  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu).
\end{align}
With these definitions, \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi, \nu) &= \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    \boxed{
    p(X\cond \nu, \npack,\bmx,\bmnu)
  = \frac{1}{p(\bmx\cond \bmnu)}
  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X_j \cond \alpha, \beta) % p(X \cond N, \bmphi, \nu)
  \, p(\bmx\cond \bmphi, \npack, \bmnu)
  \, p(\bmphi).
  }
\end{align}

@@

For large $n$, the cumulative of the Negative Binomial $p(N \cond n)$
can be approximated by the cumulative of a Gaussian with mean $n$ and
variance $2 (n - a + 1)$. Intuitively, this variance arises because the
Poisson variance appears twice, once when observing $n$ and once when
predicting $N$ given $n$.  The region around $N=n$ yields the dominant
contribution to the sum over $N$ in \refeq{gmd}. The contribution
around $N=0$ is then negligible, and we can approximate $\sum_N$ by
$\int \rmdx{N}$. The density $p(N \cond n) P(X \cond N, \npack, \bmx,
\bmnu)$ is close to a Gaussian (in the $x$ parametrization) and the
integral can be accurately approximated by Laplace's method which
requires the Hessian and the mode.
  % as each moment has to be
  % normalized by the evidence $p(\bmx \cond m)$.
The former is available in closed form and the latter has to be
determined numerically. Fortunately, Newton's method leads to a
solution in only a few steps~\cite{minka2002estimating}.
%
\fred{The reference is for max. likelihood of a $\GammaDist(\cdot)$
  but I'm sure the performance is similar for max. posterior and the
  first moments} We benefit from partial error cancelation when we
apply Laplace's method both in the numerator and
denominator~\cite{tierney_accurate_1986}.
  %
\fred{Might actually evaluate numerator and denominator at posterior
  mode to do only one instead of two optimizations;
  cf.~\cite{lindley_approximate_1980}}

@@

The central limit theorem states that for large $N$, $P(X \cond N,
\bmphi, \nu) \approx \mathcal{N}(X \cond N \, E_\bmphi[X_j \cond \nu],
N \, V_\bmphi[X_j \cond \nu])$. The expectation values have no known
closed-form solutions but we can easily and accurately approximate
them with Laplace's method.
% This is true independent of our model
assumption \refeq{gmi} that $X_j \sim \GammaDist(\alpha, \beta)$.
%
\fred{This only helps because the Laplace approx. when combined with
  large $n$ has an analytical result for the mode which saves us
  numerical optimization. But it requires extra (Laplace
  approximated?) integrals to compute $E_\bmphi[X_j \cond \nu]$ and
  $V_\bmphi[X_j \cond \nu]$. I can't get the Hessian analytically for
  $\npack$ samples. But we could use \texttt{Hesse} from
  \href{http://seal.web.cern.ch/seal/snapshot/work-packages/mathlibs/minuit/}{Minuit}
  or similar to approximate the Hessian at the mode by finite
  differences.}

@@

\subsubsection*{Large $\bmn, \bmN$,  and very large $\bmm$} \label{large-n-large-m}

>>>>

The Gamma distribution  with shape parameter $\alpha$ and rate
parameter $\beta$ is defined as
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
The support of $\GammaDist(x \cond \alpha, \beta)$ is $[0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak close to but not at the boundary. \fred{refer to plot of real
  luminosity packets in some frequency bin} In order to describe the
luminosity distribution of real packets through $\GammaDist(\alpha,
\beta)$, we have to transform the luminosities. To this end, we first
take the largest luminosity of all TARDIS packets across the entire
frequency range and increase it slightly to avoid numerical
difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.  We assume the total number of samples
across all bins is large, hence the effect of considering $\lmax$ a
constant independent of the data is negligible. We then transform the
samples $\bml \to \bmx$ as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond \nu,  \npack, \bmx,
\bmnu)$. Only in the end, we transform back to a probability density
in $L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = p(X  \cond \nu,  \npack, \bmx, \bmnu) \frac{1}{\lmax}
\end{align}
\hans{So far, $m$ appears only in your todo comment above. I would
  suggest a short discussion of this already in Section
  \ref{sec:dfnr}. We could then simply rewrite all equations below in
  terms of $m$ rather than $b$.} %
\fred{I removed $m$ whereever I saw it except below where I define it
  explicitly}
%
We observe good agreement between the best fit of a Gamma distribution
to TARDIS packet luminosities; an example is shown shown in
\reffig{gamma}.  While not obvious at first, a major numerical
simplification arises because of the $L \to X$
transformation. Rewriting the master equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond \nu,  \npack, \bml, \bmnu) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
\nu, \npack, \bml, \bmnu)
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single luminosity, $p(L \cond N=1,
\nu, \bml, \bmnu)$ is narrow in the sense that the standard deviation
relative to the mean is at the percent level. This implies that $p(L
\cond \nu, \npack, \bml, \bmnu)$ has one mode from each $N$ and all
modes are clearly separated. The transformation $L \to x$ effects a
strong overlap between the distributions $p(X \cond N, \nu, \npack, \bmx, \bmnu)$
with similar $N$, so $p(X \cond \nu, \npack, \bmx, \bmnu)$ can be well
approximated by a unimodal distribution; see \reffig{hedgehog}.
\fred{I'm not 100 \% sure but $p(X \cond \nu, \npack, \bmx, \bmnu)$ should still
  have $N \to \infty$ modes but they aren't visible}

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.48\textwidth]{hedgehog}
  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bml, \bmnu)$, right: $p(x \cond  \nu,  \npack, \bmx, \bmnu)$. I sum over $N$ in both cases. }
  \fred{I still need to pretty this up. I used the old binned approach
    with 50 samples in a bin from which I computed the mean and
    variance to plug into \refeq{gmr} but I computed the sum over $N$
    instead of the integral.}
\label{fig:hedgehog}
\end{figure}

@@

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
    X_j \sim \GammaDist(\alpha, \beta) \Rightarrow X \equiv \sum_{j=1}^N X_j \sim \GammaDist(N \alpha, \beta).
\end{align}
Let us assume that $\alpha$ and $\beta$ are functions of $\nu$ with unknown parameters $\bmphi$
\begin{align}
  \lleq{gmi}
  \alpha &= \alpha(\nu \cond \bmphi),\\
  \beta &= \beta(\nu \cond \bmphi),\\
  p(X_j \cond \bmphi, \nu) &=
  \GammaDist(X_j \cond \alpha, \beta),\\
  p(\bmX \cond N, \bmphi, \nu) &= \prod_{j=1}^N p(X_j \cond \bmphi, \nu).
\end{align}
With these definitions, \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi, \nu) &= \int \rmdx{\bmX} \delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi, \nu) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    \boxed{
    p(X\cond \nu, \npack,\bmx,\bmnu)
  = \frac{1}{p(\bmx\cond \npack, \bmnu)}
  \sum_N p(N\cond n)\, \int \rmdx{\bmphi} \GammaDist(X \cond N \alpha, \beta) % p(X \cond N, \bmphi, \nu)
  \, p(\bmx\cond \bmphi, \npack, \bmnu)
  \, p(\bmphi).
  }
\end{align}

@@

For large $n$, the cumulative of the Negative Binomial $p(N \cond n)$
can be approximated by the cumulative of a Gaussian with mean $n$ and
variance $2 (n - a + 1)$. For our purposes, the variance is $ \approx
2 n$. The intuitive reason for this value is that the Poisson variance
appears twice, once when observing $n$ and once when predicting $N$
given $n$.  The region around $N=n$ yields the dominant contribution
to the sum over $N$ in \refeq{gmd}. The contribution near $N=0$ is
then negligible, and we can approximate $\sum_N$ by $\int
\rmdx{N}$. The density $p(N \cond n) P(X \cond N, \npack, \bmx,
\bmnu)$ is close to a Gaussian (in the $x$ parametrization) and the
integral can be accurately approximated by Laplace's method which
requires the Hessian and the mode.
  % as each moment has to be
  % normalized by the evidence $p(\bmx \cond m)$.
The former is available in closed form and the latter has to be
determined numerically. Fortunately, Newton's method leads to a
solution in only a few steps~\cite{minka2002estimating}.
%
\fred{The reference is for max. likelihood of a $\GammaDist(\cdot)$
  but I'm sure the performance is similar for max. posterior and the
  first moments} We benefit from partial error cancelation when we
apply Laplace's method both in the numerator and
denominator~\cite{tierney_accurate_1986}.
  %
\fred{Might actually evaluate numerator and denominator at posterior
  mode to do only one instead of two optimizations;
  cf.~\cite{lindley_approximate_1980}}

@@

The central limit theorem states that for large $N$, $P(X \cond N,
\bmphi, \nu) \approx \mathcal{N}(X \cond N \, E_\bmphi[X_j \cond \nu],
N \, V_\bmphi[X_j \cond \nu])$. The expectation values have no known
closed-form solutions but we can easily and accurately approximate
them with Laplace's method.
%
\fred{This only helps because the Laplace approx. when combined with
  large $n$ has an analytical result for the mode which saves us
  numerical optimization. But it requires extra (Laplace
  approximated?) integrals to compute $E_\bmphi[X_j \cond \nu]$ and
  $V_\bmphi[X_j \cond \nu]$. I can't get the Hessian analytically for
  $\npack$ samples. But we could use \texttt{Hesse} from
  \href{http://seal.web.cern.ch/seal/snapshot/work-packages/mathlibs/minuit/}{Minuit}
  or similar to approximate the Hessian at the mode by finite
  differences.}

@@

\subsubsection*{Large $\bmn, \bmN$,  and large $\bmm$} \label{large-n-large-m}
